Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2024-03-26 transmission.tex
proof ch 7

First, in \cref{backcalc}, I use a backcalculation approach, based on \cref{E-inc-prev:eq:EPt}.
This approach enables estimation of incidence within (sub)populations of England.
There are several attractive features about this approach in the context here.
The backcalculation approach makes no assumptions about how transmission occurs; the assumptions have the potential to introduce bias.
It can also easily incorporate arbitrary covariates to adjust for non-response bias.
The scale of the CIS and desire to correct for non-response bias presents challenges for this methodology, especially within the limited computation of the SRS environment.

@@

In this section, I will apply the theory introduced in \cref{E-inc-prev} to estimate incidence.
I view the CIS as a series of daily, cross-sectional prevalence surveys.
This is the same approach as previously used to estimate prevalence from the CIS (see \cref{E-biology-data:sec:MRP}).

@@

The first approach, which I call the \emph{incidence-inferred} approach, is a single statistical model where incidence is inferred, with prevalence as a derived quantity.
The basis of the statistical model, in a Bayesian framework, would be as follows.
\begin{align}
    \vec{Z} &\dist F_Z \\
    P_{i,t}  &= \sum_{k=0}^{d_\text{max}-1} Z_{i,t-k} S(k+1) \\
    \rho &\dist F_\rho \\
    Y_{i,t} \mid \vec{P}, \rho &\dist \BB(n_{i,t}, P_{i,t}, \rho)
\end{align}
where $\vec{Z}$ is a vector such that $Z_{i,t}$ is the incidence proportion in stratum $i$ on day $t$, and has prior $F_Z$; $P_{i,t}$ is the prevalence in stratum $i$ on day $t$; $\rho$ is the overdispersion parameter of the beta-binomial distribution with prior $F_\rho$; $Y_{i,t}$ is the number of positive tests in stratum $i$ on day $t$; and $n_{i,t}$ is the number of tests in stratum $i$ on day $t$.
Fitting this model to a prevalence survey such as the CIS (\eg using MCMC) gives a posterior distribution of $\vec{Z}$, the target of inference.

The relationship between $P$ and $Z$ is from \cref{E-inc-prev:eq:EPt}, although dividing both sides by $\Npop$ to have the values in terms of incidence proportion and prevalence rather than incidence and number of prevalent individuals.

Specifying $F_Z$ would depend on the application of the approach, but should consider the need to provide structure to stabilise the problem (see \cref{E-intro:sec:disease-model}).
In the context of the CIS, it needs to allow borrowing information across time and/or strata; this is required because some stratum and time combinations have few positives (fewer than five per day).

@@

The incidence-inferred approach is more parsimonious in being a single step procedure.
However, it poses several pragmatic challenges which means I favour the prevalence-inferred approach.
Here I explain that further developments, beyond the scope of this thesis, would be required to use the incidence-inferred approach within the computational constraints of the SRS.

It will be convenient to define a variable equating to the posterior initially estimated.
Let $\pi_{i,t} = Z_{i,t}$ under the inference-inferred approach and $\pi_{i,t} = P_{i,t}$ is the prevalence.

The chosen approach should have the following properties.
\begin{enumerate}
    \item The prior should favour $\pi_{i,t}$ changing exponentially in $t$, \ie penalize changes from exponential growth.
        % Penalizing changes from continuous exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
        Such penalization has previously been shown to improve estimates of SARS-CoV-2 prevalence by balancing over and under-fitting~\autocite{ealesAppropriately} and is the approach used for the CIS prevalence model (see \cref{E-biology-data:sec:MRP}).
    \item Limit estimates to epidemiologically and biologically plausible values.
        That is, incidence proportion and prevalence should be in the interval $[0, 1]$.
        % While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a day which is not biologically plausible.
    \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias.
    \item Borrow information across the different covariates.
        The borrowing should penalize changes in the distribution of incidence/prevalence across strata in the same geographic region being constant.
        A constant distribution is predicted by the theoretical study of epidemics, for example the dynamics of stratified, mechanistic transmission models (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) over short periods of time with negligible behavioural and immunity changes.
    \item Be computationally efficient.
        The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see \cref{E-biology-data:sec:SRS}).
\end{enumerate}

A parsimonious model satisfying properties 1 and 2 is to specify $\logit(\pi_{i,t}) = g_\varphi(i, t)$, where $g$ is a function parameterised by $\varphi$ and the prior on $\varphi$ is specified to favour $d^2g/dt^2 = 0$, \ie favour linear growth in $\logit(\pi_{i,t}$.
The prior $F_Z$ or $F_P$ is now a prior on $\varphi$.
The $\logit$ link function both limits $\pi_{i,t}$ to the interval $[0, 1]$, and logit-linear growth approximates exponential growth for small $\pi_{i,t}$~\autocite{ealesAppropriately}.

The logit transformation also allows property 4 to be conveniently satisfied using an additive model for $g_\varphi$.
An additive model on the logit scale is multiplicative on the natural scale for the small values of $\pi_{i,t}$ which will be common.

@@

Properties 3 and 5 suggest the use of an approximate method for computing the posterior distribution.
In particular, using MCMC, sufficient covariates to adjust for the non-response bias, and a period of more than 6--8 weeks leads to a model that is computationally prohibitive ~\citePersonalComms{Koen Pouwels}.
Instead, previous MRP implementations used INLA (see \cref{E-biology-data:sec:MRP}).

INLA-based inference can only be applied to LGMs (\emph{Latent Gaussian Models}); see \cref{E-transmission:sec:INLA} for further details on INLA and LGMs.
The LGM framework requires that a transformation of $P_{i,t}$ is linear in $g_\varphi = \logit(\phi_{i,t})$.

@@

% First, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution (using the procedure described in \cref{transmission:sec:INLA:posterior}).
% For each sample and stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
% I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:observation-process} justifies, that is the estimate of the incidence is negligibly affected by approximating the distribution $\vec{P} \mid \vec{Z}$ by its mean.
% I solve this system using forward substitution, calculating the value of $Z_t$ in \cref{inc-prev:eq:forward-substitute} sequentially for $t = 1, \dots, T$.
% $t = 1$ is defined as the first day where prevalence data is used.
% \Cref{inc-prev:eq:forward-substitute} is repeated here for convenience:
% \begin{align}
% Z_t
% &= \frac{P_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
% &= P_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
% \label{transmission:eq:forward-substitute}
% \end{align}
% % \Cref{E-inc-prev:eq:EPt}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
% Dividing each side by the total number of individuals in each stratum transforms the number of prevalent individuals to prevalence and incidence to incidence proportion.

Violation of assumptions in various parts of this procedure can lead to the second term in \cref{transmission:eq:forward-substitute} to be greater than the first, giving negative incidence.
In this case, the incidence prior to day $t$ imply that the prevalence on day $t$ is higher than that predicted by the fitted regression model for this strata/day combination.
There are two sets of assumptions which are most likely to be violated.

@@

A slow increase in incidence is estimated to occur over the autumn 2020 period, peaking in late November in most regions (see \cref{transmission:fig:backcalc-regions}).
An exception is the North East and East of England regions (see \cref{transmission:fig:backcalc-regions}(C) and (D) respectively); these regions show no consistent increase or decrease in incidence during October.
All regions then decrease during November; this coincides with the national lockdown which came into force in England on 5 Nov 2020~\autocite{ifgTimeline}.

During November, incidence starts rising, with regional heterogeneity in the timing of when this occurs (see \cref{transmission:fig:backcalc-regions}).
The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocite{walkerTracking,daviesEstimated,lythgoeLineage}.
There is evidence that when incidence starts increasing is correlated with when the Alpha variant's spread (see \cref{transmission:fig:backcalc-alpha}).
A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
\begin{figure}
    \includegraphics{transmission/backcalc-regions}
    \caption[Incidence estimated using backcalculation by region]{%
        Incidence proportion for (A) all of England and (B) its component regions.
    }
    \label{transmission:fig:backcalc-regions}
\end{figure}
\begin{figure}
    \includegraphics{transmission/backcalc-alpha}
    \caption[Correlation of the emergence of Alpha and minimum incidence.]{%
        x-axis: time that S-gene target failure, a marker of the Alpha variant, is estimated to start increasing from baseline by region among CIS test results; from \textcite[table S3]{walkerTracking}.
        y-axis: posterior median and 95\% CrI of the date of minimum incidence in November or early December in each region.
        Linear regression line shown.
    }
    \label{transmission:fig:backcalc-alpha}
\end{figure}

An edge effect where a rapid rise is seen between day 1 and 2 is visible (see \cref{transmission:fig:backcalc-start-effect}).
This is due to the incorrect assumption that incidence is constant in the past.
Changing this assumption (\eg to exponential growth) modifies the estimated value on the first day, but the rest of the series is insensitive to this change (not shown).
Therefore, the first day's estimate is unreliable, and has been excluded from all other plots.
However, the rest of the series is robust to this assumption and hence reliable.
\begin{figure}
    \centering \includegraphics{transmission/backcalc-start-effect}
    \caption[Edge effects in backcalculation method]{%
        First seven days of incidence proportion estimate for the whole of England, highlighting the edge effect.
    }
    \label{transmission:fig:backcalc-start-effect}
\end{figure}

Alternative inference techniques could allow more powerful modelling approaches, if they allowed a broader set of models than LGMs.
In particular, the incidence-inferred approach could be adopted.
This could allow more interpretable priors to be used because the infection process is the process that corresponds to biological and behavioural processes.
Furthermore, the incidence proportion could be restricted to be non-negative, preventing the negative incidence seen in some strata.

@@

In common with \cref{E-SEIR:sec:mechanistic-models}, the notation in this section differs slightly from the rest of the thesis.

@@

A parsimonious model is needed to parameterize the WAIFW matrix.
In \cref{E-SEIR:sec:structured-populations}, I explained how the elements of the WAIFW matrix, $\matr{\beta}$ can be decomposed as $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'} / N_{a}$ where\todo{should I repeat these definitions? they already exist in ch 3}: $\kappa_{aa'}$ is the mean rate at which an individual in stratum $a'$ makes contact with individuals in stratum $a$; $q_{aa'}$ is the probability of transmission upon contact between an infectious individual in stratum $a'$ and a susceptible individual in stratum $a$; and $N_{a}$ is the total number of individuals in stratum $a$.
For many epidemics, where the population behaviour is fairly constant over time, a constant $\matr{\beta}$ is a good approximation.
However, in SARS-CoV-2, population behaviour changed dramatically, both spontaneously (\eg due to fear) and in response to government-imposed restrictions.
Therefore, following \textcite{birrellRealtime}, I introduce a time-dependence giving: $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'}(t) N_{a'} / N_a$; $N_a$ and $N_{a'}$ are constant because the change in the total population is negligible over the length of time considered here.

I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using the methodology of \textcite{vanleeuwenTime,vanleeuwenAugmenting}.
%; these estimates have been used for previous modelling of the SARS-CoV-2 pandemic~\autocite{vanleeuwenTime,birrellRealtime}.
% These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
This method starts with pre-pandemic baseline estimates of $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on each contact's age and location (home, work, school, leisure, transport, or other).
UKTUS asked a representative sample of the UK population about how much time they spent doing various activities (\eg shopping and spending time at home) each day.
\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
Google mobility data are based on the location of Android users over time, relative to a pre-pandemic baseline.
The mobility and school data are averaged over a week then used to down or up-weight the number of contacts in each setting, before combining contacts across settings to form $\kappa_{aa'}(t)$.
$\kappa_{aa'}(t)$ is assumed to be piecewise constant, changing only at the start of each week.
For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.

@@

\Textcite{osogamiClosed} provide a method to construct the compartments, implemented in the mapfit R package~\autocite{mapfit}. 
The resulting compartmental approximation matches the original distribution up to the first three moments.
Applying the package to the posterior mean of the distribution estimated in \cref{E-imperf-test} gives the model in \cref{SEIR:fig:full-model} with the parameter estimates in \cref{SEIR:table:ec-params} (see \cref{transmission:sec:phase-type} for details).
\begin{table}
    \centering
    \begin{tabular}{c c c c c}
        $l$ & $\omega_1$ & $\omega_{2}$ & $\omega_{3}$ & $\pi$ \\
        3 & 0.0820 & 0.126 & 0.0223 & 0.00973  \\
    \end{tabular}
    \caption{Parameter estimates for the EC distribution approximating the posterior mean of the duration distribution estimated in \cref{E-imperf-test}.}
    \label{SEIR:table:ec-params}
\end{table}
% end table of ec params

@@

    \path  (S) edge node {$\lambda$} (P0)
          (P0) edge node {$\omega_0$} (P1)
          (P1) edge node {$\omega_1$} (P2)
          (P2) edge node {$\pi \omega_{2}$} (P3)
               edge [out=-45,in=-135] node[below] {$(1 - \pi) \omega_{2}$} (P4)
          (P3) edge node {$\omega_{3}$} (P4);

@@

A diagram of the model for a single stratum is shown in figure \cref{SEIR:fig:full-model}.
The full system of differential equations for each region is below.
\begin{equation}
\begin{aligned}
    \vec{\lambda}(t) &= \beta(t) (\matr{\kappa}(t) \circ \matr{q} \circ \matr{N}) ((\vec{i_1}(t) + \vec{i_2}(t))) \\
    \frac{d\vec{s}(t)}{dt} &= -\vec{\lambda}(t) \circ \vec{s}(t) \\
    \frac{d\vec{e_1}(t)}{dt} &= \vec{\lambda}(t) \circ \vec{s}(t) - 2\sigma \vec{e_1}(t) \\
    \frac{d\vec{e_2}(t)}{dt} &= 2\sigma \vec{e_1}(t) - 2\sigma \vec{e_2}(t) \\
    \frac{d\vec{i_1}(t)}{dt} &= 2\sigma \vec{e_2}(t) - 2\gamma \vec{i_1}(t) \\
    \frac{d\vec{i_2}(t)}{dt} &= 2\gamma \vec{i_1}(t) - 2\gamma \vec{i_2}(t) \\
    \frac{d\vec{r}(t)}{dt} &= 2\gamma \vec{i_2}(t) \\
    \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \omega_0 \vec{p_0}(t) \\
    \frac{d\vec{p_1}(t)}{dt} &= \omega_0 \vec{p_0}(t) - \omega_1 \vec{p_1}(t) \\
    \frac{d\vec{p_2}(t)}{dt} &= \omega_1 \vec{p_1}(t) - \omega_{2} \vec{p_2}(t) \\
    \frac{d\vec{p_3}(t)}{dt} &= \pi \omega_{2} \vec{p_2}(t) - \omega_{3} \vec{p_3}(t)
\end{aligned}
\label{SEIR:eq:fullODEs}
\end{equation}
where $\circ$ indicates element-wise multiplication and $\matr{N}$ is a matrix with $(\matr{N})_{aa'} = N_{a'} / N_{a}$.
See \cref{SEIR:eq:fullODEs}

@@

>>>>

First, in \cref{backcalc}, I use a backcalculation approach, based on \cref{E-inc-prev:eq:EPt}.
This approach enables estimation of incidence within (sub)populations of England.
There are several attractive features about this approach in the context here.
The backcalculation approach makes few assumptions about how transmission occurs; the assumptions have the potential to introduce bias.
It can also easily incorporate arbitrary covariates to adjust for non-response bias.
The scale of the CIS and desire to correct for non-response bias presents challenges for this methodology, especially within the limited computation of the SRS environment.

@@

In this section, I apply the theory introduced in \cref{E-inc-prev} to estimate incidence.
I view the CIS as a series of daily, cross-sectional prevalence surveys.
This is the same approach as previously used to estimate prevalence from the CIS (see \cref{E-biology-data:sec:MRP}).

@@

The first approach, which I call the \emph{incidence-inferred} approach, is a single statistical model where incidence is inferred, with prevalence as a derived quantity.
The basis of the statistical model, in a Bayesian framework, would be as follows.
\begin{align}
    \vec{Z} &\dist F_Z \\
    P_{i,t}  &= \sum_{k=0}^{d_\text{max}-1} Z_{i,t-k} S(k+1) \\
    \rho &\dist F_\rho \\
    Y_{i,t} \mid \vec{P}, \rho &\dist \BB(n_{i,t}, P_{i,t}, \rho)
\end{align}
where $\vec{Z}$ is a vector such that $Z_{i,t}$ is the incidence proportion in stratum $i$ on day $t$, with a joint prior $F_Z$; $P_{i,t}$ is the prevalence in stratum $i$ on day $t$; $\rho$ is the overdispersion parameter of the beta-binomial distribution with prior $F_\rho$; $Y_{i,t}$ is the number of positive tests in stratum $i$ on day $t$; and $n_{i,t}$ is the number of tests in stratum $i$ on day $t$.
Fitting this model to a prevalence survey such as the CIS (\eg using MCMC) gives a posterior distribution of $\vec{Z}$, the target of inference.

The relationship between $P$ and $Z$ is from \cref{E-inc-prev:eq:EPt}, although dividing both sides by the stratum's population (assumed constant) to have the values in terms of incidence proportion and prevalence rather than incidence and number of prevalent individuals.

Specifying $F_Z$ would depend on the application of the approach, but should consider the need to provide structure to stabilise the problem (see \cref{E-intro:sec:disease-model}).
In the context of the CIS, it needs to allow borrowing information across time and/or strata (see \cref{E-biology-data:sec:MRP}).

@@

The incidence-inferred approach is more parsimonious in being a single step procedure.
However, it poses several pragmatic challenges which means I favour the prevalence-inferred approach.

It will be convenient to define a variable equating to the quantity initially estimated.
Let $\pi_{i,t} = Z_{i,t}$ under the inference-inferred approach and $\pi_{i,t} = P_{i,t}$ is the prevalence.

The chosen approach should have the following properties.
\begin{enumerate}
    \item The prior should favour $\pi_{i,t}$ changing exponentially in $t$, \ie penalize changes from exponential growth.
        % Penalizing changes from continuous exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
        Such penalization has previously been shown to improve estimates of SARS-CoV-2 prevalence by balancing over and under-fitting~\autocite{ealesAppropriately} and is the approach used for the CIS prevalence model (see \cref{E-biology-data:sec:MRP}).
    \item Limit estimates to epidemiologically and biologically plausible values.
        That is, incidence proportion and prevalence should be in the interval $[0, 1]$.
        % While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a day which is not biologically plausible.
    \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias.
    \item Borrow information across the different covariates.
        The borrowing should penalize changes in the distribution of incidence/prevalence across strata in the same geographic region.
        A constant distribution is predicted by the theoretical study of epidemics, for example the dynamics of stratified, mechanistic transmission models (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) over short periods of time with negligible behavioural and immunity changes.
    \item Be computationally efficient.
        The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see \cref{E-biology-data:sec:SRS}).
\end{enumerate}

A parsimonious model satisfying properties 1 and 2 is to specify $\logit(\pi_{i,t}) = g_\varphi(i, t)$, where $g$ is a function parameterised by $\varphi$ and the prior on $\varphi$ is specified to favour $d^2g/dt^2 = 0$, \ie favour linear growth in $\logit(\pi_{i,t})$.
The prior $F_Z$ or $F_P$ is now a prior on $\varphi$.
The $\logit$ link function both limits $\pi_{i,t}$ to the interval $[0, 1]$, and logit-linear growth approximates exponential growth for small $\pi_{i,t}$~\autocite{ealesAppropriately}.

The logit transformation also allows property 4 to be conveniently satisfied using an additive model for $g_\varphi$.
An additive model on the logit scale is approximately multiplicative on the natural scale for small values of $\pi_{i,t}$ (which will be common).

@@

Properties 3 and 5 suggest the use of an approximate method for computing the posterior distribution.
In particular, using MCMC, sufficient covariates to adjust for the non-response bias, over a period of more than 6--8 weeks leads to a model that is computationally prohibitive ~\citePersonalComms{Koen Pouwels}.
Instead, previous MRP implementations for the CIS used INLA (see \cref{E-biology-data:sec:MRP}).

INLA-based inference can only be applied to LGMs (\emph{Latent Gaussian Models}); see \cref{E-transmission:sec:INLA} for further details on INLA and LGMs.
The LGM framework requires that a transformation of $P_{i,t}$ is linear in a vector of $g_{\varphi}(i, t)$s combining all $i$s and $t$s.

@@

% First, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution (using the procedure described in \cref{transmission:sec:INLA:posterior}).
% For each sample and stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
% I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:observation-process} justifies, that is the estimate of the incidence is negligibly affected by approximating the distribution $\vec{P} \mid \vec{Z}$ by its mean.
% I solve this system using forward substitution, calculating the value of $Z_t$ in \cref{inc-prev:eq:forward-substitute} sequentially for $t = 1, \dots, T$.
% $t = 1$ is defined as the first day where prevalence data are used.
% \Cref{inc-prev:eq:forward-substitute} is repeated here for convenience:
% \begin{align}
% Z_t
% &= \frac{P_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
% &= P_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
% \label{transmission:eq:forward-substitute}
% \end{align}
% % \Cref{E-inc-prev:eq:EPt}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
% Dividing each side by the total number of individuals in each stratum transforms the number of prevalent individuals to prevalence and incidence to incidence proportion.

Violation of assumptions in various parts of this procedure can lead to the second term in \cref{transmission:eq:forward-substitute} to be greater than the first, giving negative incidence.
In this case, the incidence prior to day $t$ imply that the prevalence on day $t$ is higher than that predicted by the fitted regression model for this strata/day combination.

A slow increase in incidence is estimated to occur over the autumn 2020 period, peaking in October in most regions (see \cref{transmission:fig:backcalc-regions}).
An exception is the North East and East of England regions (see \cref{transmission:fig:backcalc-regions}(C) and (D) respectively); these regions show no consistent increase or decrease in incidence during October.
All regions then decrease during November; this coincides with the national lockdown which came into force in England on 5 Nov 2020~\autocite{ifgTimeline}.

During November, incidence starts rising, with regional heterogeneity in the timing of when this occurs (see \cref{transmission:fig:backcalc-regions}).
The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocite{walkerTracking,daviesEstimated,lythgoeLineage}.
There is evidence that when incidence starts increasing is correlated with the timing of the Alpha variant's spread (see \cref{transmission:fig:backcalc-alpha}).
A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
\begin{figure}
    \includegraphics{transmission/backcalc-regions}
    \caption[Incidence estimated using backcalculation by region]{%
        Incidence proportion for (A) all of England and (B) its component regions.
    }
    \label{transmission:fig:backcalc-regions}
\end{figure}
\begin{figure}
    \includegraphics{transmission/backcalc-alpha}
    \caption[Correlation of the emergence of Alpha and minimum incidence.]{%
        x-axis: time that S-gene target failure, a marker of the Alpha variant, is estimated to start increasing from baseline by region among CIS test results; from \textcite[table S3]{walkerTracking}.
        y-axis: posterior median and 95\% CrI of the date of minimum incidence in November or early December in each region.
        Linear regression line shown.
    }
    \label{transmission:fig:backcalc-alpha}
\end{figure}

An edge effect, a rapid rise in incidence between days one and two, is visible (see \cref{transmission:fig:backcalc-start-effect}).
This is due to the incorrect assumption that incidence is constant in the past.
Changing this assumption (\eg to exponential growth) modifies the estimated value on the first day, but the rest of the series is insensitive to this change (not shown).
Therefore, the first day's estimate is unreliable, and has been excluded from all other results.
However, the rest of the series is robust to this assumption and hence reliable.
\begin{figure}
    \centering \includegraphics{transmission/backcalc-start-effect}
    \caption[Edge effects in backcalculation method]{%
        First seven days of incidence proportion estimate for the whole of England, highlighting the edge effect.
    }
    \label{transmission:fig:backcalc-start-effect}
\end{figure}

The issues with negative incidence are likely caused by issues with modelling assumptions.
There are two sets of assumptions which are most likely to be violated.

@@

Alternative inference techniques could allow more powerful modelling approaches, if they allowed a broader set of models than LGMs.
In particular, the incidence-inferred approach could be adopted.
This could allow more interpretable priors to be used because the infection process is the process that corresponds to biological and behavioural processes.
Furthermore, it would prevent the issues with negative incidence.

@@

In common with \cref{E-SEIR:sec:mechanistic-models}, the notational conventions in this section differs slightly from the rest of the thesis.

@@

A parsimonious model is needed to parameterize the WAIFW matrix.
In \cref{E-SEIR:sec:structured-populations}, I explained how the elements of the WAIFW matrix, $\matr{\beta}$ can be decomposed as $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'} / N_{a}$ where\todo{should I repeat these definitions? they already exist in ch 3}: $\kappa_{aa'}$ is the mean rate at which an individual in stratum $a'$ makes contact with individuals in stratum $a$; $q_{aa'}$ is the probability of transmission upon contact between an infectious individual in stratum $a'$ and a susceptible individual in stratum $a$; and $N_{a}$ is the total number of individuals in stratum $a$.
For many epidemics, where the population behaviour is fairly constant over time, a constant $\matr{\beta}$ is a good approximation.
However, in SARS-CoV-2, population behaviour changed dramatically, both spontaneously (\eg due to fear) and in response to government-imposed restrictions.
Therefore, following \textcite{birrellRealtime}, I introduce a time-dependence giving: $\beta_{aa'}(t) = \kappa_{aa'}(t) q_{aa'}(t) N_{a'} / N_a$; $N_a$ and $N_{a'}$ are constant because the change in the total population is negligible over the length of time considered here.

I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using the methodology of \textcite{vanleeuwenTime,vanleeuwenAugmenting}.
%; these estimates have been used for previous modelling of the SARS-CoV-2 pandemic~\autocite{vanleeuwenTime,birrellRealtime}.
% These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
This method starts with pre-pandemic baseline estimates of $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on each contact's age and location (home, work, school, leisure, transport, or other).
UKTUS asked a representative sample of the UK population about how much time they spent doing various activities (\eg shopping and spending time at home) each day.
\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
Google mobility data are based on the location of Android users over time, relative to a pre-pandemic baseline.
The mobility and school data are averaged over a week down or up-weight the number of contacts in each setting, before combining contacts across settings to form $\kappa_{aa'}(t)$.
$\kappa_{aa'}(t)$ is assumed to be piecewise constant, changing only at the start of each week.
For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.

@@

\Textcite{osogamiClosed} provide a method to construct the compartments, implemented in the mapfit R package~\autocite{mapfit}. 
The resulting compartmental approximation matches the original distribution up to the first three moments.
Applying the package to the posterior mean of the distribution estimated in \cref{E-imperf-test} gives the model in \cref{SEIR:fig:full-model} with the parameter estimates in \cref{SEIR:table:ec-params} (see \cref{transmission:sec:phase-type} for details).
\begin{table}
    \centering
    \begin{tabular}{c c c c c}
        $l$ & $\omega_1$ & $\omega_{2}$ & $\omega_{3}$ & $\chi$ \\
        3 & 0.0820 & 0.126 & 0.0223 & 0.00973  \\
    \end{tabular}
    \caption{Parameter estimates for the phase-type distribution with the same first three moments as the posterior mean of the duration distribution estimated in \cref{E-imperf-test}.}
    \label{SEIR:table:ec-params}
\end{table}
% end table of ec params

@@

    \path  (S) edge node {$\lambda$} (P0)
          (P0) edge node {$\omega_0$} (P1)
          (P1) edge node {$\omega_1$} (P2)
          (P2) edge node {$\chi \omega_{2}$} (P3)
               edge [out=-45,in=-135] node[below] {$(1 - \chi) \omega_{2}$} (P4)
          (P3) edge node {$\omega_{3}$} (P4);

@@

A diagram of the model for a single stratum is shown in figure \cref{SEIR:fig:full-model}.
The full system of differential equations for each region is below.
\begin{equation}
\begin{aligned}
    \vec{\lambda}(t) &= \beta(t) (\matr{\kappa}(t) \circ \matr{q} \circ \matr{N}) ((\vec{i_1}(t) + \vec{i_2}(t))) \\
    \frac{d\vec{s}(t)}{dt} &= -\vec{\lambda}(t) \circ \vec{s}(t) \\
    \frac{d\vec{e_1}(t)}{dt} &= \vec{\lambda}(t) \circ \vec{s}(t) - 2\sigma \vec{e_1}(t) \\
    \frac{d\vec{e_2}(t)}{dt} &= 2\sigma \vec{e_1}(t) - 2\sigma \vec{e_2}(t) \\
    \frac{d\vec{i_1}(t)}{dt} &= 2\sigma \vec{e_2}(t) - 2\gamma \vec{i_1}(t) \\
    \frac{d\vec{i_2}(t)}{dt} &= 2\gamma \vec{i_1}(t) - 2\gamma \vec{i_2}(t) \\
    \frac{d\vec{r}(t)}{dt} &= 2\gamma \vec{i_2}(t) \\
    \frac{d\vec{p_0}(t)}{dt} &= \vec{\lambda} \circ \vec{s}(t) - \omega_0 \vec{p_0}(t) \\
    \frac{d\vec{p_1}(t)}{dt} &= \omega_0 \vec{p_0}(t) - \omega_1 \vec{p_1}(t) \\
    \frac{d\vec{p_2}(t)}{dt} &= \omega_1 \vec{p_1}(t) - \omega_{2} \vec{p_2}(t) \\
    \frac{d\vec{p_3}(t)}{dt} &= \chi \omega_{2} \vec{p_2}(t) - \omega_{3} \vec{p_3}(t)
\end{aligned}
\label{SEIR:eq:fullODEs}
\end{equation}
where $\circ$ indicates element-wise multiplication and $\matr{N}$ is a matrix with $(\matr{N})_{aa'} = N_{a'} / N_{a}$.
See \cref{SEIR:eq:fullODEs}

@@

Both reparameterizations rely on an assumption that the epidemic is in a steady-state exponential growth phase at time 0 (see \cref{E-SEIR:sec:SIR}). During such an exponential growth phase, all components of the $\vec{e}$, $\vec{i}$ and $\vec{p}$ compartments grow at the same growth rate, $\psi$.
That is:
\begin{equation}
\begin{aligned}
\frac{d\vec{e_1}(t)}{dt} = \psi \vec{e_1}(t) \\
\frac{d\vec{e_2}(t)}{dt} = \psi \vec{e_2}(t) \\
\frac{d\vec{i_1}(t)}{dt} = \psi \vec{i_1}(t) \\
\frac{d\vec{i_2}(t)}{dt} = \psi \vec{i_2}(t) \\
\frac{d\vec{p_0}(t)}{dt} = \psi \vec{p_0}(t) \\
\frac{d\vec{p_1}(t)}{dt} = \psi \vec{p_1}(t) \\
\frac{d\vec{p_2}(t)}{dt} = \psi \vec{p_2}(t) \\
\frac{d\vec{p_3}(t)}{dt} = \psi \vec{p_3}(t).
\end{aligned}
\label{SEIR:eq:exponential-growth}
\end{equation}
This follows directly from the ODEs if $\vec{s}$ is constant.

The growth rate $\psi$ is easily identifiable from the data, and therefore a parameterization in terms of $\psi$ should be more computationally efficient.
If this is the initial exponential phase, when $\vec{s} \approx 1$, then the basic reproduction number $\R$ for an SEIR model is linked to this growth rate by~\autocite{birrellBayesian,wearingAppropriate}:
\begin{align}
    R_{\psi} = \frac{\psi \left( \frac{\psi}{2\sigma} + 1 \right)^2}{\gamma \left(1 - \frac{1}{\left(\frac{\psi}{2 \gamma} + 1 \right)^2} \right)} \label{SEIR:eq:rtoR}.
\end{align}
